Type-Safe Modular Parsing

### 2017-01-13 paper/sections/Inheritance.tex
Fix refs in Section 2 & 3

\subsection{The Mismatch Between OO ASTs and Grammars}
Object-oriented ASTs are ASTs encoded via standard OO techniques,
such as the {\sc Composite} or {\sc Interpreter} patterns~\cite{gamma94design}. Such ASTs
fix the operations required in the AST, but allow the easy addition of new
language constructs. This is in contrast with ASTs modelled using
functional programming algebraic datatypes or OO style Visitors, where
adding new operations is easy, but adding new language constructs is hard.
The AST in Figure~\ref{} provides a simple example of an OO AST. The same
AST could be encoded using algebraic datatypes (or alternatively Visitors)
as:

@@

Unfortunatelly, when parsing OO style ASTs, there are difficulties with
the two dimensions of extensibility. On the one hand it is naturally hard to
add new operations to the AST, as usual in the OO style. On the hand
grammars are not modularly extensible, thus the addition of new
language constructs is problematic. Therefore,
it is perhaps not a suprise that parser generators that also generate
AST code (auch as ANTLR~\cite{}), normally generate code based on
visitors. Clearly this seems a better choice since generating code
based on the {\sc Interpreter} or {\sc Composite} pattern would
simply be too inflexible, due to the dual limitation on extensibility.

@@

Unfortunately, this fails to parse some expressions like \inlinecode{"1 + x"}, which are obviously valid in the new grammar.
The reason is that \inlinecode{pAdd} makes two recursive calls to parse sub-expressions, by using \inlinecode{pExpr}, which
covers both cases in the old grammar. But the newly added case \inlinecode{pVar} is not observed by the recursive \inlinecode{pExpr},
hence the parser does not work as expected. It is possible to build the correct parser by replacing the recursive call in \inlinecode{pAdd} with \inlinecode{pVarExpr}.
However, modification on existing code sacrifices separate compilation, as mentioned by the Expression Problem.

@@

\subsection{Multiple Extensions and Independent Extensibility}
A nice feature of Scala is its support for a form of multiple inheritance.
Scala has a linearized-style multiple inheritance for traits~\cite{}.
This can be very helpful when composing several languages, which
may have been independently developed. Suppose now
we want to compose the parsers for expressions from pre-defined
languages \inlinecode{LanguageA} and \inlinecode{LanguageB} using
alternative. The new parser can be built by inheriting both parsers
at the same time:

@@

\huang{added explanation of the `super' keyword}

@@

\huang{added a concrete example}

>>>>

\subsection{The Mismatch Between OO ASTs and Grammars}
Object-oriented ASTs are ASTs encoded via standard OO techniques,
such as the {\sc Composite} or {\sc Interpreter} patterns~\cite{gamma1995design}. Such ASTs
fix the operations required in the AST, but allow the easy addition of new
language constructs. This is in contrast with ASTs modelled using
functional programming algebraic datatypes or OO style Visitors, where
adding new operations is easy, but adding new language constructs is hard.
In Section~\ref{subsec:packratparsing} we provide a simple example of OO ASTs. The same abstract syntax could be encoded using algebraic datatypes (or alternatively Visitors)
as:

@@

Unfortunately, when parsing OO style ASTs, there are difficulties with
the two dimensions of extensibility. On the one hand it is naturally hard to
add new operations to the AST, as usual in the OO style. On the hand
grammars are not modularly extensible, thus the addition of new
language constructs is problematic. Therefore,
it is perhaps not a surprise that parser generators that also generate
AST code (such as ANTLR~\cite{parr2013definitive}), normally generate code based on
visitors. Clearly this seems a better choice since generating code
based on the {\sc Interpreter} or {\sc Composite} pattern would
simply be too inflexible, due to the dual limitation on extensibility.

@@

Unfortunately, this fails to parse some expressions like \inlinecode{"1 + x"}, which are obviously valid in the new grammar.
The reason is that \inlinecode{pAdd} makes two recursive calls to parse sub-expressions, by using \inlinecode{pExpr}, which
covers both cases in the old grammar. But the newly added case \inlinecode{pVar} is not observed by the recursive \inlinecode{pExpr},
hence the parser does not work as expected. It is possible to build the correct parser by replacing the recursive call in \inlinecode{pAdd} with \inlinecode{pVarExpr}.
However, modification on existing code sacrifices separate compilation, and hence breaks semantic modularity.

@@

\subsection{Multiple Extensions and Independent Extensibility}
A nice feature of Scala is its support for a form of multiple inheritance.
Scala has a linearized-style multiple inheritance for traits~\cite{}\huang{what's the proper citation here?}.
This can be very helpful when composing several languages, which
may have been independently developed. Suppose now
we want to compose the parsers for expressions from pre-defined
languages \inlinecode{LanguageA} and \inlinecode{LanguageB} using
alternative. The new parser can be built by inheriting both parsers
at the same time:

@@

@@

### 2017-01-13 paper/sections/PackratParsing.tex
Fix refs in Section 2 & 3

This section discusses the algorithmic challenges introduced by
modular parsing and argues that Packrat parser combinators
are suitable to address those challenges. The algorithm challenges
are important because they rule out various common techniques
used by non-modular code using parsing combinators.
To avoid pitfalls related to those algortihmic challenges,
we propose a the following methodology:

@@

\subsection{Algorithmic Challenges of Modularity}\label{subsec:challenges}
At a first look, parser combinators are very suitable for modular parsing, because of two reasons. Firstly, they are naturally modular. The manner of using them is to write small parsers and use combinators to composed them together. The construction procedure is explicit and fully controlled by the programmer. Secondly, each parser combinator is represented by a piece of code, and also are the parsers it takes. Thus in a statically typed programming language they can be statically type-checked.
Unfortunately many parser combinators have important limitations.
In particular several parser combinators,
including the famous Parsec~\cite{Leijen2001} library, require
programmers to manually do \textit{left-recursion elimination}, \textit{longest match composition}, and
require significant amounts of \textit{backtracking}. All of those are
problematic in a modular setting.

@@

After left-recursion elimination, the structure of grammar is changed,
as well as its corresponding parser. In a modular setting, it is
possible but unnecessarily compilcated to analyse the grammar and
rewrite it when doing extension. Anticipating that every non-terminal
has left-recursive rules is helpful for extensibility but overkill,
since it is inconvenient and introduces extra complexity for
representation of grammar and implementation of parser.

@@

In this case, reordering the alternatives ensures that
the \emph{longest match} is picked among the possible results. However, manual reording for the longest match is inconvenient, and worst still, it is essentially non-modular. When the grammar is extended with new rules, programmers are supposed to \emph{manually} adjust
the order of parsers, which requires rewritting previously written code.

@@

\subsection{Packrat Parsing}\label{subsec:packratparsing}
Fortunately some more advanced parsing techniques such as Packrat
parsing~\cite{Ford2002} have been developed to address limitations of
simple parser combinators. Packrat parsers use memoization to record
the result of applying each parser at each position of the input, so
that repeated computation is eliminated. Moreover, Packrat parsing, in
theory, supports both direct and indirect left-recursion~\cite{warth2008}. All of these properties are very suitable for modularity, thus we decided to use Packrat parsers as the underlying parsing technique in for modular parsing.
\begin{comment}
It is worth mentioning that the choice of parser combinators will not
affect the other parts of our library. One can choose other parser
combinators like Parsec, in cases that the performance and supporting
of left-recursion are not major concerns. A different library can even build a new
\name with fancy features or higher efficiency.
\end{comment}
Scala has a standard parser combinator library\footnote{https://github.com/scala/scala-parser-combinators}
\cite{moors2008parser} which implements Packrat parsers.
The library provides a number of parser combinators, including the longeset match alternative combinator.
Below we present an example to illustrate Scala parsers.

\paragraph{Code Demonstration}
For more concise demonstration, we assume that all the Scala code in the rest of this paper are in the object \inlinecode{Code}, as shown in Figure~\ref{fig:papercode}. It extends traits \inlinecode{StandardTokenParsers} and \inlinecode{PackratParsers} from the Scala parser combinator library. Futhermore, we will use \lstinline{Parser} as a type synonym for \lstinline{PackratParser} and a generic \inlinecode{parse} function for testing.

@@

In the trait \lstinline{ExprParser}, \lstinline{lexical} is used for lexing. \lstinline{pLit} parses an integer for the literal case.
\lstinline{pAdd} handles the addition case and creates an object of \lstinline{Add}. It parses two sub-expressions by calling \lstinline{pExpr}
recursively. Finally \lstinline{pExpr} composes \lstinline{pLit} and \lstinline{pAdd} using the longest match alternative combinator \inlinecode{|||}.
Table~\ref{tab:packrat} gives a closer look at part of the API used in our work.

\begin{table}[t]
\begin{tabular}{l}
\hline
\begin{lstlisting}
def ~[U](q: => Parser[U]): Parser[~[T, U]]
\end{lstlisting} \\
\hspace{.2in}- A parser combinator for sequential composition. \\
\hline
\begin{lstlisting}
def ^^[U](f: (T) => U): Parser[U]
\end{lstlisting} \\
\hspace{.2in}- A parser combinator for function application. \\
\hline
\begin{lstlisting}
def ^^^[U](v: => U): Parser[U]
\end{lstlisting} \\
\hspace{.2in}- A parser combinator that changes a successful result into the specified value. \\
\hline
\begin{lstlisting}
def <~[U](q: => Parser[U]): Parser[T]
\end{lstlisting} \\
\hspace{.2in}- A parser combinator for sequential composition which keeps only the left result. \\
\hline
\begin{lstlisting}
def ~>[U](q: => Parser[U]): Parser[U]
\end{lstlisting} \\
\hspace{.2in}- A parser combinator for sequential composition which keeps only the right result. \\
\hline
\begin{lstlisting}
def repsep[T](p: => Parser[T], q: => Parser[Any]): Parser[List[T]]
\end{lstlisting} \\
\hspace{.2in}- A parser generator for interleaved repetitions. \\
\hline
\begin{lstlisting}
def ident: Parser[String]
\end{lstlisting} \\
\hspace{.2in}- A parser which matches an identifier. \\
\hline
\begin{lstlisting}
def numericLit: Parser[String]
\end{lstlisting} \\
\hspace{.2in}- A parser which matches a numeric literal. \\
\hline
\begin{lstlisting}
def |[U >: T](q: => Parser[U]): Parser[U]
\end{lstlisting} \\
\hspace{.2in}- A parser combinator for alternative composition. \\
\hline
\begin{lstlisting}
def |||[U >: T](q0: => Parser[U]): Parser[U]
\end{lstlisting} \\
\hspace{.2in}- A parser combinator for alternative with longest match composition. \\
\hline \\
\end{tabular}
\caption{Part of Scala Parser API.\bruno{We should double-check that this table lists the
  combinators used in the paper.}}\label{tab:packrat}
\end{table}

>>>>

This section discusses the algorithmic challenges introduced by
modular parsing and argues that Packrat parser combinators
are suitable to address those challenges. The algorithm challenges
are important because they rule out various common techniques
used by non-modular code using parsing combinators.
To avoid pitfalls related to those algorithmic challenges,
we propose the following methodology:

@@

\subsection{Algorithmic Challenges of Modularity}\label{subsec:challenges}
At a first look, parser combinators are very suitable for modular parsing, because of two reasons. Firstly, they are naturally modular. The manner of using them is to write small parsers and use combinators to composed them together. The construction procedure is explicit and fully controlled by the programmer. Secondly, each parser combinator is represented by a piece of code, also are the parsers it takes. Thus in a statically typed programming language they can be statically type-checked.
Unfortunately many parser combinators have important limitations.
In particular, several parser combinators including the famous Parsec~\cite{Leijen2001} library, require
programmers to manually do \textit{left-recursion elimination}, \textit{longest match composition}, and
require significant amounts of \textit{backtracking}. All of those are
problematic in a modular setting.

@@

After left-recursion elimination, the structure of grammar is changed,
as well as its corresponding parser. In a modular setting, it is
possible but unnecessarily complicated to analyse the grammar and
rewrite it when doing extension. Anticipating that every non-terminal
has left-recursive rules is helpful for extensibility but overkill,
since it is inconvenient and introduces extra complexity for
representation of grammar and implementation of parser.

@@

In this case, reordering the alternatives ensures that
the \emph{longest match} is picked among the possible results. However, manual reordering for the longest match is inconvenient, and worst still, it is essentially non-modular. When the grammar is extended with new rules, programmers are supposed to \emph{manually} adjust
the order of parsers, which requires rewriting previously written code.

@@

\subsection{Packrat Parsing}\label{subsec:packratparsing}
Fortunately, some more advanced parsing techniques such as Packrat
parsing~\cite{Ford2002} have been developed to address limitations of
simple parser combinators. Packrat parsers use memoization to record
the result of applying each parser at each position of the input, so
that repeated computation is eliminated. Moreover, Packrat parsing, in
theory, supports both direct and indirect left-recursion~\cite{warth2008}. All of these properties are very suitable for modularity, thus we decided to use Packrat parsers as the underlying parsing technique in for modular parsing.
\begin{comment}
It is worth mentioning that the choice of parser combinators will not
affect the other parts of our library. One can choose other parser
combinators like Parsec, in cases that the performance and supporting
of left-recursion are not major concerns. A different library can even build a new
\name with fancy features or higher efficiency.
\end{comment}
Scala has a standard parser combinator library\footnote{https://github.com/scala/scala-parser-combinators}
\cite{moors2008parser} which implements Packrat parsers.
The library provides a number of parser combinators, including the longeset match alternative combinator.
Below we present an example to illustrate Scala parsers.

\paragraph{Code Demonstration}
For more concise demonstration, we assume that all the Scala code in the rest of this paper are in the object \inlinecode{Code}, as shown in Figure~\ref{fig:papercode}. It extends traits \inlinecode{StandardTokenParsers} and \inlinecode{PackratParsers} from the Scala parser combinator library. Furthermore, we will use \lstinline{Parser} as a type synonym for \lstinline{PackratParser} and a generic \inlinecode{parse} function for testing.

@@

In the trait \lstinline{ExprParser}, \lstinline{lexical} is used for lexing. \lstinline{pLit} parses an integer for the literal case.
\lstinline{pAdd} handles the addition case and creates an object of \lstinline{Add}. It parses two sub-expressions by calling \lstinline{pExpr}
recursively. Finally \lstinline{pExpr} composes \lstinline{pLit} and \lstinline{pAdd} using the longest match alternative combinator \inlinecode{|||}.
Table~\ref{tab:packrat} gives a closer look at the parser combinators used in our work.

\begin{table}[t]
\begin{tabular}{l}
\hline
\begin{lstlisting}
def ~[U](q: => Parser[U]): Parser[~[T, U]]
\end{lstlisting} \\
\hspace{.2in}- A parser combinator for sequential composition. \\
\hline
\begin{lstlisting}
def ^^[U](f: (T) => U): Parser[U]
\end{lstlisting} \\
\hspace{.2in}- A parser combinator for function application. \\
\hline
\begin{lstlisting}
def ^^^[U](v: => U): Parser[U]
\end{lstlisting} \\
\hspace{.2in}- A parser combinator that changes a successful result into the specified value. \\
\hline
\begin{lstlisting}
def <~[U](q: => Parser[U]): Parser[T]
\end{lstlisting} \\
\hspace{.2in}- A parser combinator for sequential composition which keeps only the left result. \\
\hline
\begin{lstlisting}
def ~>[U](q: => Parser[U]): Parser[U]
\end{lstlisting} \\
\hspace{.2in}- A parser combinator for sequential composition which keeps only the right result. \\
\hline
\begin{lstlisting}
def repsep[T](p: => Parser[T], q: => Parser[Any]): Parser[List[T]]
\end{lstlisting} \\
\hspace{.2in}- A parser generator for interleaved repetitions. \\
\hline
\begin{lstlisting}
def ident: Parser[String]
\end{lstlisting} \\
\hspace{.2in}- A parser which matches an identifier. \\
\hline
\begin{lstlisting}
def numericLit: Parser[String]
\end{lstlisting} \\
\hspace{.2in}- A parser which matches a numeric literal. \\
\hline
\begin{lstlisting}
def |[U >: T](q: => Parser[U]): Parser[U]
\end{lstlisting} \\
\hspace{.2in}- A parser combinator for alternative composition. \\
\hline
\begin{lstlisting}
def |||[U >: T](q0: => Parser[U]): Parser[U]
\end{lstlisting} \\
\hspace{.2in}- A parser combinator for alternative with longest match composition. \\
\hline \\
\end{tabular}
\caption{Combinators used in our work, from the Scala standard parser combinator library.}\label{tab:packrat}
\end{table}
